---
title: "Disaggregation Workflow (Noah/Patrick)"
output: pdf_document
header-includes:
  - |
    ```{=latex}
    \expandafter\def\csname ver@subfig.sty\endcsname{}
    \usepackage{graphicx,subcaption}
    \makeatletter\@ifundefined{subfloat}{\newcommand{\subfloat}[2][need a sub-caption]{\subcaptionbox{#1}{#2} }}{\renewcommand{\subfloat}[2][need a sub-caption]{\subcaptionbox{#1}{#2} }}\makeatother
    ```
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
knitr::opts_chunk$set(cache.path = "cache/rmd/")
if (any(commandArgs() == "mdToTex")) {
  knitr::knit_hooks$set(plot = knitr::hook_plot_tex)
}
```


```{r libraries}
library(ggplot2)
library(dplyr)
library(tidyr)
library(Matrix)
library(aghq)
library(disaggMultiMap)
library(terra)
library(sf)
library(data.table)
```


```{r helpers}
plot_priors_posteriors_manual <- function(fit_object) {
  opt  <- fit_object$aghq_model$optresults
  mode <- opt$mode
  hess <- opt$hessian
  cov_mat <- tryCatch(solve(hess), error = function(e) MASS::ginv(hess))
  par_names <- names(mode)

  data_list <- fit_object$obj$env$data

  req_vars <- c(
    "prior_rho_min", "prior_rho_prob", "prior_sigma_max",
    "prior_sigma_prob", "prior_iideffect_sd_max", "prior_iideffect_sd_prob"
  )
  missing <- setdiff(req_vars, names(data_list))
  if (length(missing) > 0) {
    stop(paste("Error: Missing required prior data in fit_object:", paste(missing, collapse = ", ")))
  }

  rho_min  <- as.numeric(data_list$prior_rho_min)
  rho_prob <- as.numeric(data_list$prior_rho_prob)
  sig_max  <- as.numeric(data_list$prior_sigma_max)
  sig_prob <- as.numeric(data_list$prior_sigma_prob)
  tau_max  <- as.numeric(data_list$prior_iideffect_sd_max)
  tau_prob <- as.numeric(data_list$prior_iideffect_sd_prob)

  get_post_dens <- function(param_name, transform_fun, jacobian_fun, x_range_sd = 4) {
    idx <- which(par_names == param_name)
    if (length(idx) == 0) return(NULL)
    mu <- mode[idx]
    sd <- sqrt(cov_mat[idx, idx])
    x_est <- seq(mu - x_range_sd * sd, mu + x_range_sd * sd, length.out = 1000)
    d_est <- dnorm(x_est, mean = mu, sd = sd)
    x_nat <- transform_fun(x_est)
    jac <- jacobian_fun(x_est)
    d_nat <- d_est * (1 / abs(jac))
    if (any(abs(x_nat) < (0.1 * max(abs(x_nat))))) {
      x_nat <- c(0, x_nat)
      d_nat <- c(0, d_nat)
    }
    data.frame(x = x_nat, y = d_nat, type = "Posterior")
  }

  plot_list <- list()

  if ("log_rho" %in% par_names) {
    df_post <- get_post_dens("log_rho", exp, exp)
    lam_rho <- -log(rho_prob) * rho_min
    x_grid <- df_post$x[df_post$x >= 0]
    y_prior <- (lam_rho / x_grid^2) * exp(-lam_rho / x_grid)
    df_prior <- data.frame(x = x_grid, y = y_prior, type = "Prior")
    plot_list[["rho"]] <- rbind(df_post %>% filter(x >= 0), df_prior) %>%
      mutate(param = "Range (Rho) [m]")
  }

  if ("log_sigma" %in% par_names) {
    df_post <- get_post_dens("log_sigma", exp, exp)
    lam_sig <- -log(sig_prob) / sig_max
    x_grid <- df_post$x
    y_prior <- dexp(x_grid, rate = lam_sig)
    df_prior <- data.frame(x = x_grid, y = y_prior, type = "Prior")
    plot_list[["sigma"]] <- rbind(df_post, df_prior) %>%
      mutate(param = "Spatial SD (Sigma)")
  }

  if ("iideffect_log_tau" %in% par_names) {
    df_post <- get_post_dens("iideffect_log_tau", exp, exp)
    lam_tau <- -log(tau_prob) / tau_max
    x_grid <- df_post$x
    y_prior <- dexp(x_grid, rate = lam_tau)
    df_prior <- data.frame(x = x_grid, y = y_prior, type = "Prior")
    plot_list[["tau"]] <- rbind(df_post, df_prior) %>%
      mutate(param = "NB Dispersion (Tau)")
  }

  beta_idx <- grep("intercept|slope|beta", par_names, ignore.case = TRUE)
  exclude <- which(par_names %in% c("log_rho", "log_sigma", "iideffect_log_tau"))
  beta_idx <- setdiff(beta_idx, exclude)

  for (i in beta_idx) {
    pname <- par_names[i]
    df_post <- get_post_dens(pname, identity, function(x) 1)
    x_grid <- df_post$x
    y_prior <- dnorm(x_grid, mean = 0, sd = 100)
    dname <- tools::toTitleCase(gsub("_", " ", pname))
    df_prior <- data.frame(x = x_grid, y = y_prior, type = "Prior")
    plot_list[[pname]] <- rbind(df_post, df_prior) %>% mutate(param = dname)
  }

  full_df <- bind_rows(plot_list)
  hyper_names <- c("Range (Rho) [m]", "Spatial SD (Sigma)", "NB Dispersion (Tau)")

  p1 <- NULL
  df_hyper <- full_df %>% filter(param %in% hyper_names)
  if (nrow(df_hyper) > 0) {
    p1 <- ggplot(df_hyper, aes(x = x, y = y)) +
      geom_line(aes(color = type, linetype = type), linewidth = 1.1) +
      scale_color_manual(values = c("Posterior" = "navy", "Prior" = "firebrick")) +
      scale_linetype_manual(values = c("Posterior" = "solid", "Prior" = "dashed")) +
      facet_wrap(~param, scales = "free") +
      theme_bw() +
      labs(title = "Hyperparameters", subtitle = "PC Priors (Red) vs Posterior (Blue)", x = "Value", y = "Density")
  }

  p2 <- NULL
  df_fixed <- full_df %>% filter(!param %in% hyper_names)
  if (nrow(df_fixed) > 0) {
    p2 <- ggplot(df_fixed, aes(x = x, y = y)) +
      geom_line(aes(color = type, linetype = type), linewidth = 1.1) +
      scale_color_manual(values = c("Posterior" = "navy", "Prior" = "firebrick")) +
      scale_linetype_manual(values = c("Posterior" = "solid", "Prior" = "dashed")) +
      facet_wrap(~param, scales = "free") +
      theme_bw() +
      labs(title = "Fixed Effects", subtitle = "Priors (Red) vs Posterior (Blue)", x = "Value", y = "Density")
  }

  list(hyper = p1, fixed = p2, df = full_df)
}

plot_sf <- function(sf_df, field_name) {
  ggplot2::ggplot(sf_df) +
    ggplot2::geom_sf(ggplot2::aes(fill = .data[[field_name]])) +
    ggplot2::scale_fill_viridis_c(limits = range(sf_df[[field_name]], na.rm = TRUE))
}
```


```{r data-setup, echo=FALSE}
res_choice <- "1km"  # "1km" or "100m"
data_dir <- "../data"
options(timeout = 3600)

Syears <- c(2003, 2006, 2010, 2013, 2016, 2021, 2024)
Surl <- paste0(
  "https://gisco-services.ec.europa.eu/distribution/v2/nuts/gpkg/NUTS_RG_10M_",
  Syears, "_3035.gpkg"
)
names(Surl) <- Syears

Sfiles <- Pmisc::downloadIfOld(Surl, path = data_dir)
nutsAll <- mapply(terra::vect, x = Sfiles, SIMPLIFY = FALSE)
names(nutsAll) <- Syears

nuts <- lapply(nutsAll, function(x) x[x$CNTR_CODE %in% c("NL", "BE") & x$LEVL_CODE == 3, ])

nutsYear <- list(
  "2018" = nuts[["2016"]],
  "2019" = nuts[["2016"]],
  "2020" = nuts[["2021"]],
  "2021" = nuts[["2021"]],
  "2022" = nuts[["2021"]],
  "2023" = nuts[["2024"]]
)

mort <- eurostat::get_eurostat("demo_r_magec3", type = "code", cache_dir = data_dir)
mort$year <- as.numeric(gsub("-.*", "", mort$TIME_PERIOD))

popDf <- expand.grid(
  year = names(nutsYear),
  country = c("NLD", "BEL"),
  age = c("00", "01")
)

wp_base <- "https://data.worldpop.org/GIS/AgeSex_structures/Global_2015_2030/R2025A"
if (res_choice == "1km") {
  popDf$url <- paste0(
    wp_base, "/", popDf$year, "/", popDf$country,
    "/v1/1km_ua/constrained/", tolower(popDf$country),
    "_f_", popDf$age, "_", popDf$year, "_CN_1km_R2025A_UA_v1.tif"
  )
} else if (res_choice == "100m") {
  popDf$url <- paste0(
    wp_base, "/", popDf$year, "/", popDf$country,
    "/v1/100m/constrained/", tolower(popDf$country),
    "_f_", popDf$age, "_", popDf$year, "_CN_100m_R2025A_v1.tif"
  )
} else {
  stop("res_choice must be '100m' or '1km'")
}

popDf$files <- Pmisc::downloadIfOld(popDf$url, path = data_dir)

popDfSplit <- split(popDf, popDf$country)
popSplit <- mapply(function(xx) terra::rast(xx$files), popDfSplit, SIMPLIFY = FALSE)

pop <- terra::mosaic(
  terra::tapp(popSplit[[1]], popDfSplit[[1]]$year, sum),
  terra::tapp(popSplit[[2]], popDfSplit[[2]]$year, sum)
)

popT1 <- terra::project(pop, terra::crs(nuts[[1]]))
pop_file <- file.path(data_dir, paste0("pop_", res_choice, ".tif"))
popT <- terra::writeRaster(popT1, pop_file, overwrite = TRUE)
popT <- terra::rast(pop_file)

mort1 <- as.data.frame(mort[mort$unit == "NR" & mort$age == "Y_LT5" & mort$sex == "F", ])
for (D in names(nutsYear)) {
  mortHere <- mort1[mort1$year == as.integer(D), ]
  nutsYear[[D]]$deaths <- mortHere[match(nutsYear[[D]]$NUTS_ID, mortHere$geo), "values"]
}
crop_region <- terra::aggregate(nutsYear[["2018"]], dissolve = TRUE)
```



```{r maps-overview, fig.cap="Population and boundary overview maps for NL/BE.", fig.subcap=c("Population raster with NUTS level-3 boundaries", "Country context map with selected NUTS-year outlines"), fig.show='hold', out.width='0.44\\linewidth', fig.width=6, fig.height=5}
invisible(lapply(nutsYear, function(xx) quantile(xx$deaths)))

pCol <- mapmisc::colourScale(breaks = c(0, 0.1, 0.5, 1, 2, 20), style = "fixed")
mapmisc::map.new(popT)
plot(popT[[1]], breaks = pCol$breaks, col = pCol$col, legend = FALSE, maxcell = 1e7, add = TRUE)
plot(nuts[[6]], add = TRUE)
mapmisc::legendBreaks("topleft", pCol, inset = 0.1, bty = "n")

data("worldMap", package = "mapmisc")
world <- terra::unwrap(worldMap)
worldT <- terra::project(world, terra::crs(nuts[[1]]))

nuts3 <- nutsYear[c(1, 3, 6)]
Scol <- c("lightblue", "green", "blue")
Slwd <- seq(2, 0.8, len = length(nuts3))^2
Slty <- rep(1, length(nuts3))
mapmisc::map.new(worldT[grep("Belgium|Nether", worldT$NAME), ], buffer = 10 * 1000)
plot(popT[[1]], breaks = pCol$breaks, col = pCol$col, legend = FALSE, maxcell = 1e7, add = TRUE)
for (D in seq_along(nuts3)) {
  plot(nuts3[[D]], lwd = Slwd[D], add = TRUE, border = Scol[D], lty = Slty[D])
}
legend("topleft", lwd = 3, col = Scol, legend = names(nuts3), bty = "n", inset = 0.1)
```



```{r build-disagg-data, cache=TRUE}
poly_list <- lapply(nutsYear, sf::st_as_sf)
offset_list <- setNames(lapply(names(popT), function(nm) popT[[nm]]), sub("^X", "", names(popT)))

limits <- sf::st_bbox(poly_list[[1]])
hypotenuse <- sqrt((limits$xmax - limits$xmin)^2 + (limits$ymax - limits$ymin)^2)
maxedge <- hypotenuse / 10

prep_time <- system.time({
  dis_data_mmap <- disaggMultiMap::prepare_data_mmap(
    polygon_shapefile_list = poly_list,
    aggregation_rasters_list = offset_list,
    mesh_args = list(
      max.edge = c(maxedge / 5, maxedge),
      cutoff = 0.2 * maxedge,
      offset = c(hypotenuse / 4, hypotenuse / 2),
      resolution = 250
    ),
    response_var = "deaths",
    id_var = "NUTS_ID",
    na_action = TRUE,
    verbose = TRUE
  )
})["elapsed"]

prep_time
```

```{r disagg-data-diagnostics, fig.cap = 'disagg diagnostics', fig.subcap = c('base', 'fancy'), out.width = "0.45\\textwidth", fig.height=6, fig.width=4}
mapmisc::map.new(terra::vect(poly_list[[1]]), buffer = 100 * 1000)
plot(terra::vect(poly_list[[1]]), add = TRUE)
points(dis_data_mmap$mesh$loc[, 1:2], pch = 15, cex = 0.4)

#print(paste0(nrow(dis_data_mmap$mesh$graph$tv), " basis functions"))
plot(dis_data_mmap)
```

## Priors

```{r priors, echo=TRUE}

my_priors <- disaggMultiMap::get_priors(dis_data_mmap)
my_priors$prior_rho_min <- 25000
my_priors$prior_rho_prob <- 0.5
my_priors$prior_iideffect_sd_max <- 0.1
my_priors$prior_iideffect_sd_prob <- 0.5
```



```{r fit-pooled, cache=TRUE}
fm_nb_k1_pooled_priros <- disaggMultiMap::disag_model_mmap(
  data = dis_data_mmap,
  aghq_k = 2,
  family = "negbinomial",
  link = "log",
  iid = TRUE,
  verbose = TRUE,
  priors = my_priors,
  engine = "AGHQ",
  time_varying_betas = FALSE
)
```

```{r predictions, eval=FALSE}
pred_k1_pooled_priors <- predict(fm_nb_k1_pooled_priros, verbose = TRUE)
pred_mask <- terra::mask(
  pred_k1_pooled_priors$mean_prediction$prediction$time_1, 
  crop_region)
predMaskWrap = terra::wrap(pred_mask)  
```


```{r post-sampling, eval=exists('fm_nb_k1_pooled_priros'), cache=TRUE}
samps <- aghq::sample_marginal(fm_nb_k1_pooled_priros$aghq_model, 500)
mats <- disaggMultiMap:::get_predict_matrices(
  data = fm_nb_k1_pooled_priros$data,
  new_data = NULL,
  expected_cov_names = tryCatch(
    fm_nb_k1_pooled_priros$model_setup$coef_meta$cov_names,
   error = function(e) NULL),
  time_varying_betas = isTRUE(fm_nb_k1_pooled_priros$model_setup$time_varying_betas)
)

rfield <- mats$A %*% samps$samps
res_k1 <- rast(
  as.data.frame(cbind(
    mats$coords,
    med = apply(rfield, 1, median),
    exc = apply(rfield, 1, function(xx) mean(xx > 0))
  )),
  crs = terra::crs(worldT)
)
pred_mask <- terra::mask(res_k1, crop_region)
predMaskWrap = terra::wrap(pred_mask)  
```

aa


```{r pooled-maps, fig.cap='stuff', out.width='0.5\\textwidth'}
pred_mask = terra::unwrap(predMaskWrap)
pred_col <- mapmisc::colourScale(
  pred_mask[["med"]],
  dec = 2,
  breaks = 8,
  style = "equal",
  col = "Spectral",
  rev = TRUE
)

mapmisc::map.new(pred_mask)
plot(pred_mask$med, col = pred_col$col, breaks = pred_col$breaks, legend = FALSE, add = TRUE)
mapmisc::legendBreaks("topleft", pred_col, bty = "n")
```




```{r prior-posterior-export, eval=exists('fm_nb_k1_pooled_priros')}
plots_priors <- plot_priors_posteriors_manual(fm_nb_k1_pooled_priros)
par_plot_split <- split(plots_priors$df, plots_priors$df$param)
par_plot_split2 <- lapply(
  lapply(par_plot_split, data.table::as.data.table),
  data.table::dcast,
  formula = param + x ~ type,
  value.var = "y"
)
```


```{r plotPriorPost, eval=exists('par_plot_split2'), fig.cap = 'Priors and posteriors', fig.subcap = c('nb','range','spatial sd'), dev='pdf', out.width='0.31\\textwidth', fig.height=3, fig.width=3}
par(mar = c(2, 4, 0.5, 0))
matplot(
  par_plot_split2$NB$x,
  par_plot_split2$NB[, c("Prior", "Posterior")],
  lty = c(2, 1), type = "l", col = "black", bty = "l",
  ylab = "dens", yaxs = "i", xaxs = "i", xlim = c(0, 0.42)
)

matplot(
  par_plot_split2$Range$x / 1000,
  par_plot_split2$Range[, c("Prior", "Posterior")] * 1000,
  lty = c(2, 1), type = "l", col = "black", bty = "l",
  ylab = "dens", yaxs = "i", xaxs = "i", xlim = c(0, 75)
)

matplot(
  par_plot_split2$Spatial$x,
  par_plot_split2$Spatial[, c("Prior", "Posterior")],
  lty = c(2, 1), type = "l", col = "black", bty = "l",
  ylab = "dens", yaxs = "i", xaxs = "i", xlim = c(0, 2.75)
)
```

## Export Maps


```{r plotSetup, echo=FALSE}
pred_mask = terra::unwrap(predMaskWrap)

dir.create("cache", showWarnings = FALSE)
borders <- terra::vect(c(
  geodata::gadm("NLD", level = 0, path = "cache"),
  geodata::gadm("FRA", level = 0, path = "cache"),
  geodata::gadm("BEL", level = 0, path = "cache"),
  geodata::gadm("CHE", level = 0, path = "cache"),
  geodata::gadm("AUT", level = 0, path = "cache"),
  geodata::gadm("ITA", level = 0, path = "cache"),
  geodata::gadm("LIE", level = 0, path = "cache"),
  geodata::gadm("DEU", level = 0, path = "cache"),
  geodata::gadm("GBR", level = 0, path = "cache"),
  geodata::gadm("LUX", level = 0, path = "cache")
))
bordersT <- terra::project(borders, terra::crs(worldT))

pred_col_med <- mapmisc::colourScale(
  exp(pred_mask$med),
  dec = 1,
  breaks = 8,
  style = "equal",
  col = "Reds",
  transform = "log",
  rev = FALSE
)
pred_col_exc <- mapmisc::colourScale(
  breaks = c(0, 0.2, 0.5, 0.8, 0.95, 1),
  style = "fixed",
  col = "RdYlGn",
  rev = TRUE
)

theExt = apply(as.matrix(ext(pred_mask)),1,diff)  
```

```{r export-maps, fig.cap="Disaggregation map outputs for the NL/BE region.", fig.subcap=c("Posterior median surface (exp scale)", "Exceedance probability (Pr > 0)", "Boundaries", "Population 2018", "Deaths 2018", "Basis vertices"), out.width='0.31\\linewidth', fig.width=4, fig.height=4*theExt[2]/theExt[1]}



mapmisc::map.new(pred_mask, bg = "lightblue")
plot(pred_mask$med, col = pred_col_med$col, breaks = log(pred_col_med$breaks), legend = FALSE, add = TRUE, colNA = par("bg"))
plot(bordersT, col = c("#FFFFFFFF", "#FFFFFF00")[1 + grepl("Netherlands|Belgium", bordersT$COUNTRY)], add = TRUE)
mapmisc::legendBreaks("topleft", pred_col_med)

mapmisc::map.new(pred_mask, bg = "lightblue")
plot(pred_mask$exc, col = pred_col_exc$col, breaks = pred_col_exc$breaks, legend = FALSE, add = TRUE, colNA = par("bg"))
plot(bordersT, col = c("#FFFFFFFF", "#FFFFFF00")[1 + grepl("Netherlands|Belgium", bordersT$COUNTRY)], add = TRUE)
mapmisc::legendBreaks("topleft", pred_col_exc)

mapmisc::map.new(pred_mask, bg = "#FFFFFF00")
plot(bordersT, col = "white", add = TRUE)
for (D in seq_along(nuts3)) {
  plot(nuts3[[D]], lwd = Slwd[D], add = TRUE, border = Scol[D], lty = Slty[D])
}
#plot(bordersT, col = c("#FFFFFFFF", "#FFFFFF00")[1 + grepl("Netherlands|Belgium", bordersT$COUNTRY)], add = TRUE)
legend("topleft", lwd = 3, col = Scol, legend = names(nuts3), bty='n')

mapmisc::map.new(pred_mask, bg = "lightblue")
plot(bordersT, col = "white", add = TRUE)
plot(popT[["X2018"]], breaks = pCol$breaks, col = pCol$col, legend = FALSE, maxcell = 1e7, add = TRUE)
plot(bordersT, col = c("#FFFFFFFF", "#FFFFFF00")[1 + grepl("Netherlands|Belgium", bordersT$COUNTRY)], add = TRUE)
mapmisc::legendBreaks("topleft", pCol)

deathsCol = mapmisc::colourScale(nutsYear[['2018']]$deaths,
   breaks = c(0,1,5,10,15, 20, 25, 35), style='fixed',col='Reds')

mapmisc::map.new(pred_mask, bg = "lightblue")
plot(bordersT, col = "white", add = TRUE, border='white')
plot(nutsYear[['2018']], add=TRUE, col = deathsCol$plot)
#plot(bordersT, col = c("#FFFFFFFF", "#FFFFFF00")[1 + grepl("Netherlands|Belgium", bordersT$COUNTRY)], add = TRUE, lty=0)
mapmisc::legendBreaks("topleft", deathsCol)


vecpoints = terra::vect(dis_data_mmap$mesh$loc[, 1:2], crs = crs(bordersT))
mapmisc::map.new(vecpoints, buffer = c(0, 100*1000))
plot(bordersT, add = TRUE, col='white', border='grey')
points(vecpoints, pch = 15, cex = 0.4)

```




cc

```{r fit-timevarying, cache=TRUE, eval=FALSE}
# does'nt work
fm_nb_k1 <- disaggMultiMap::disag_model_mmap(
  data = dis_data_mmap,
  aghq_k = 1,
  family = "negbinomial",
  link = "log",
  iid = TRUE,
  verbose = TRUE,
  priors = my_priors,
  engine = "AGHQ",
  time_varying_betas = TRUE
)
```

```{r pred-timevarying, cache=TRUE, eval=exists('fm_nb_k1')}
pred_k1 <- predict(fm_nb_k1, verbose = TRUE)
summary(fm_nb_k1)
```



I would like to replicate the analysis in the file `disagg.Rmd` to estimate the risk of under-5 mortality in Mancheters, UK for 2008 to 2012 using death counts from the ONS at the LSOA level and population from worldpop.  I know the 2008 and 2009 data are on one set of boundaries and 2010 onwards are on a different set.  Please write an RMD file to download and assemble the data in the correct form.  Also include code similar to disagg.Rmd to run the model, but don't run it yet (put eval=FALSE in the chunks)
