---
title: "Manchester Under-5 Mortality Disaggregation (2008-2012)"
geometry: margin=1in
header-includes:
  - |
    ```{=latex}
    \expandafter\def\csname ver@subfig.sty\endcsname{}
    \usepackage{graphicx,subcaption}
    \makeatletter\@ifundefined{subfloat}{\newcommand{\subfloat}[2][need a sub-caption]{\subcaptionbox{#1}{#2} }}{\renewcommand{\subfloat}[2][need a sub-caption]{\subcaptionbox{#1}{#2} }}\makeatother
    ```
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  cache = FALSE,fig.height=4, fig.width=5
)
options(timeout = 3600)
options(download.file.method = "libcurl")
knitr::opts_chunk$set(cache.path = "cache/man/")
knitr::opts_chunk$set(fig.path = "figure/man/")

knitr::knit_hooks$set(plot = knitr::hook_plot_tex)
# Use GEOS for polygon ops; avoids s2 loop-duplication topology errors on UK boundaries.
sf::sf_use_s2(FALSE)
```


```{r install-disaggmultimap, echo=TRUE}
if (!requireNamespace("disaggMultiMap", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
  }
  remotes::install_github("nripstein/disaggMultiMap")
}

if (!requireNamespace("mapmisc", quietly = TRUE)) {
  install.packages("mapmisc")
}
```

```{r libraries, include=FALSE}
library(dplyr)
library(tidyr)
library(data.table)
library(sf)
library(terra)
library(jsonlite)
library(readr)
library(readxl)
library(mapmisc)
library(disaggMultiMap)
library(aghq)
```


```{r config, echo=FALSE}
years <- 2008:2012
# LAD codes to include.
# Manchester = E08000003, Salford = E08000006
target_lad_codes <- c(
  "E08000001", "E08000002", "E08000003", "E08000004", "E08000005",
  "E08000006", "E08000007", "E08000008", "E08000009", "E08000010"
)
uk_grid_crs <- 27700

# Set to "100m" if you need that resolution from WorldPop.
worldpop_res <- "1km"
# Raster preprocessing controls for memory/performance.
offset_buffer_m <- 5000
offset_agg_fact <- 2

# Local data directory.
data_dir <- "../data/manchester_u5"
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
terra_tmp <- file.path(data_dir, "terra_tmp")
dir.create(terra_tmp, recursive = TRUE, showWarnings = FALSE)
terra::terraOptions(tempdir = terra_tmp, memfrac = 0.6, progress = 1)

# ONS deaths archive (covers 2001-2019; includes the 2008-2012 period).
ons_deaths_zip_url <- paste0(
  "https://www.ons.gov.uk/file?uri=",
  "%2Fpeoplepopulationandcommunity%2Fbirthsdeathsandmarriages%2Fdeaths%2Fadhocs%2F",
  "12626deathsbylowerlayersuperoutputarealsoaenglandandwalesmidyear2001to2019%2F",
  "deathsbylsoamidyear01to19.zip"
)

# ONS Open Geography / ArcGIS data sources.
url_lsoa11_bdy_geojson <- paste0(
  "https://opendata.arcgis.com/api/v3/datasets/",
  "02e8d336d6804fbeabe6c972e5a27b16_0/downloads/data?format=geojson&spatialRefId=4326"
)

url_lsoa01_bdy_geojson <- paste0(
  "https://opendata.arcgis.com/api/v3/datasets/",
  "98b573a6f3ea483e97f7e08771624316_0/downloads/data?format=geojson&spatialRefId=4326"
)

# LSOA(2001) -> LSOA(2011) -> LAD(2011) best-fit lookup (ArcGIS FeatureServer).
url_lsoa01_to_lsoa11_service <- paste0(
  "https://services1.arcgis.com/ESMARspQHYMw9BZ9/ArcGIS/rest/services/",
  "LSOA01_LSOA11_LAD11_EW_LU_ddfe1cd1c2784c9b991cded95bc915a9/FeatureServer/0/query"
)
```

## Helpers

```{r helpers, echo=FALSE}
download_if_missing <- function(url, destfile) {
  needs_download <- !file.exists(destfile) || isTRUE(file.info(destfile)$size == 0)
  if (!needs_download) {
    return(destfile)
  }

  dir.create(dirname(destfile), recursive = TRUE, showWarnings = FALSE)
  tmp <- paste0(destfile, ".part")
  methods <- c("libcurl", "curl", "auto")
  last_err <- NULL

  for (attempt in 1:4) {
    for (m in methods) {
      ok <- tryCatch({
        if (file.exists(tmp)) file.remove(tmp)
        suppressWarnings(
          download.file(
            url = url,
            destfile = tmp,
            mode = "wb",
            quiet = FALSE,
            method = m,
            headers = c("User-Agent" = "R/disagg notebook")
          )
        )
        file.exists(tmp) && file.info(tmp)$size > 0
      }, error = function(e) {
        last_err <<- e$message
        FALSE
      })

      if (isTRUE(ok)) {
        if (file.exists(destfile)) file.remove(destfile)
        file.rename(tmp, destfile)
        return(destfile)
      }
    }
    Sys.sleep(2 ^ attempt)
  }

  stop("Failed to download: ", url, if (!is.null(last_err)) paste0(" (last error: ", last_err, ")"))
}

pick_col <- function(nms, patterns, required = TRUE) {
  idx <- Reduce(`|`, lapply(patterns, function(p) grepl(p, nms, ignore.case = TRUE)))
  out <- nms[idx]
  if (length(out) == 0 && required) {
    stop("Could not find required column. Patterns tried: ", paste(patterns, collapse = ", "))
  }
  if (length(out) == 0) return(NA_character_)
  out[[1]]
}

norm_code <- function(x) {
  toupper(trimws(as.character(x)))
}

read_ons_deaths_table <- function(zip_path, years_keep = 2008:2012) {
  zip_tbl <- unzip(zip_path, list = TRUE)
  inner_name <- zip_tbl$Name[grepl("\\.(csv|xlsx?)$", zip_tbl$Name, ignore.case = TRUE)][1]
  if (is.na(inner_name)) {
    stop("No CSV/XLSX death table found inside ", zip_path)
  }

  extracted <- unzip(zip_path, files = inner_name, exdir = dirname(zip_path), overwrite = TRUE)

  parse_death_table <- function(raw) {
    nms <- names(raw)

    year_col <- pick_col(nms, c("^year$", "^mid-?year$", "mid.?year", "time"))
    deaths_u1_col <- pick_col(nms, c("^<\\s*1$", "under\\s*1", "^0\\s*[-to]{0,3}\\s*0$"), required = FALSE)
    deaths_u4_col <- pick_col(nms, c("^01\\s*[-to]{0,3}\\s*04$", "^1\\s*[-to]{0,3}\\s*4$", "01-04"), required = FALSE)
    deaths_5_9_col <- pick_col(nms, c("^05\\s*[-to]{0,3}\\s*09$", "^5\\s*[-to]{0,3}\\s*9$", "05-09"), required = FALSE)
    deaths_10_14_col <- pick_col(nms, c("^10\\s*[-to]{0,3}\\s*14$", "10-14"), required = FALSE)
    deaths_15_19_col <- pick_col(nms, c("^15\\s*[-to]{0,3}\\s*19$", "15-19"), required = FALSE)

    lsoa01_col <- pick_col(nms, c("lsoa.*2001.*(code|cd)", "lsoa01(cd|code)"), required = FALSE)
    lsoa11_col <- pick_col(nms, c("lsoa.*2011.*(code|cd)", "lsoa11(cd|code)"), required = FALSE)
    lsoa_any_col <- pick_col(nms, c("^lsoa(cd|code)$", "lsoa.*(code|cd)$"), required = FALSE)

    year_raw <- as.character(raw[[year_col]])
    year_parsed <- suppressWarnings(as.integer(year_raw))
    if (all(is.na(year_parsed))) {
      year_parsed <- suppressWarnings(as.integer(sub(".*(20[0-9]{2}).*", "\\1", year_raw)))
    }
    year_parsed <- data.table::nafill(year_parsed, type = "locf")

    if (any(is.na(c(deaths_u1_col, deaths_u4_col, deaths_5_9_col, deaths_10_14_col, deaths_15_19_col)))) {
      stop("Could not find one or more 0-19 deaths columns: <1, 01-04, 05-09, 10-14, 15-19.")
    }

    out <- raw %>%
      mutate(
        year = year_parsed,
        deaths = as.numeric(.data[[deaths_u1_col]]) +
          as.numeric(.data[[deaths_u4_col]]) +
          as.numeric(.data[[deaths_5_9_col]]) +
          as.numeric(.data[[deaths_10_14_col]]) +
          as.numeric(.data[[deaths_15_19_col]])
      ) %>%
      filter(year %in% years_keep)

    out <- out %>%
      mutate(
        lsoa01cd = if (!is.na(lsoa01_col)) norm_code(.data[[lsoa01_col]]) else NA_character_,
        lsoa11cd = if (!is.na(lsoa11_col)) norm_code(.data[[lsoa11_col]]) else NA_character_
      )

    if (!is.na(lsoa_any_col)) {
      out <- out %>%
        mutate(
          lsoa01cd = if_else(is.na(lsoa01cd), norm_code(.data[[lsoa_any_col]]), lsoa01cd),
          lsoa11cd = if_else(is.na(lsoa11cd), norm_code(.data[[lsoa_any_col]]), lsoa11cd)
        )
    }

    out %>%
      transmute(year, lsoa01cd, lsoa11cd, deaths) %>%
      filter(!is.na(deaths))
  }

  read_xlsx_sheet <- function(sheet_name) {
    probe <- readxl::read_excel(extracted, sheet = sheet_name, col_names = FALSE, n_max = 80)
    first_col <- trimws(tolower(as.character(probe[[1]])))
    header_row <- which(first_col == "mid-year")[1]
    if (is.na(header_row)) header_row <- 1
    raw <- as.data.table(readxl::read_excel(extracted, sheet = sheet_name, skip = header_row - 1))
    parse_death_table(raw)
  }

  if (grepl("\\.csv$", extracted, ignore.case = TRUE)) {
    out <- parse_death_table(data.table::fread(extracted))
  } else {
    sheets <- readxl::excel_sheets(extracted)
    persons_sheet <- sheets[grepl("deaths\\s*-\\s*persons|persons", sheets, ignore.case = TRUE)][1]
    if (is.na(persons_sheet)) persons_sheet <- sheets[[1]]
    out <- read_xlsx_sheet(persons_sheet)
  }

  out
}

fetch_arcgis_attrs <- function(base_query_url, out_fields, where = "1=1", page_size = 100) {
  fetch_json <- function(url) {
    tmp <- tempfile(fileext = ".json")
    on.exit(unlink(tmp), add = TRUE)
    methods <- c(getOption("download.file.method", "auto"), "curl", "auto")
    methods <- unique(methods)
    last_err <- NULL

    for (attempt in 1:4) {
      for (m in methods) {
        ok <- tryCatch({
          if (file.exists(tmp)) file.remove(tmp)
          suppressWarnings(
            download.file(
              url = url,
              destfile = tmp,
              mode = "wb",
              quiet = TRUE,
              method = m,
              headers = c("User-Agent" = "R/disagg notebook")
            )
          )
          file.exists(tmp) && file.info(tmp)$size > 0
        }, error = function(e) {
          last_err <<- e$message
          FALSE
        })
        if (isTRUE(ok)) {
          return(jsonlite::fromJSON(tmp, simplifyVector = FALSE))
        }
      }
      Sys.sleep(2 ^ attempt)
    }
    stop("ArcGIS download failed for URL: ", substr(url, 1, 140), if (!is.null(last_err)) paste0(" ... (last error: ", last_err, ")"))
  }

  attrs_to_dt <- function(features) {
    attrs <- lapply(features, function(f) f$attributes)
    attrs <- attrs[vapply(attrs, function(x) !is.null(x) && length(x) > 0, logical(1))]
    if (length(attrs) == 0) return(NULL)
    rbindlist(lapply(attrs, as.list), fill = TRUE, use.names = TRUE)
  }

  ids_url <- paste0(
    base_query_url,
    "?where=", URLencode(where, reserved = TRUE),
    "&returnIdsOnly=true&f=pjson"
  )
  ids_res <- fetch_json(ids_url)

  if (!is.null(ids_res$error)) {
    err_msg <- if (!is.null(ids_res$error$message)) ids_res$error$message else "Unknown error"
    stop("ArcGIS ID query failed: ", err_msg)
  }

  rows <- list()
  object_ids <- ids_res$objectIds

  if (!is.null(object_ids) && length(object_ids) > 0) {
    chunks <- split(object_ids, ceiling(seq_along(object_ids) / page_size))
    for (i in seq_along(chunks)) {
      oid_str <- paste(chunks[[i]], collapse = ",")
      q <- paste0(
        base_query_url,
        "?objectIds=", URLencode(oid_str, reserved = TRUE),
        "&outFields=", URLencode(paste(out_fields, collapse = ","), reserved = TRUE),
        "&f=pjson&returnGeometry=false"
      )
      res <- fetch_json(q)
      if (!is.null(res$error)) {
        err_msg <- if (!is.null(res$error$message)) res$error$message else "Unknown error"
        stop("ArcGIS chunk query failed (chunk ", i, "): ", err_msg)
      }
      if (!is.null(res$features) && length(res$features) > 0) {
        dt <- attrs_to_dt(res$features)
        if (!is.null(dt)) rows[[length(rows) + 1L]] <- dt
      }
    }
  } else {
    q <- paste0(
      base_query_url,
      "?where=", URLencode(where, reserved = TRUE),
      "&outFields=", URLencode(paste(out_fields, collapse = ","), reserved = TRUE),
      "&f=pjson&returnGeometry=false"
    )
    res <- fetch_json(q)
    if (!is.null(res$error)) {
      err_msg <- if (!is.null(res$error$message)) res$error$message else "Unknown error"
      stop("ArcGIS query failed: ", err_msg)
    }
    if (!is.null(res$features) && length(res$features) > 0) {
      dt <- attrs_to_dt(res$features)
      if (!is.null(dt)) rows[[length(rows) + 1L]] <- dt
    }
  }

  if (length(rows) == 0) {
    return(data.table::as.data.table(setNames(replicate(length(out_fields), character(), simplify = FALSE), out_fields)))
  }
  rbindlist(rows, fill = TRUE, use.names = TRUE)
}

pick_first_existing_col <- function(df, candidates) {
  hit <- candidates[candidates %in% names(df)]
  if (length(hit) == 0) {
    stop("None of these columns were found: ", paste(candidates, collapse = ", "))
  }
  hit[[1]]
}
```

## Download Core Inputs

```{r download-inputs}
# Boundaries
lsoa11_geojson <- file.path(data_dir, "lsoa11_bgc_ew.geojson")
lsoa01_geojson <- file.path(data_dir, "lsoa01_bgc_ew.geojson")
download_if_missing(url_lsoa11_bdy_geojson, lsoa11_geojson)
download_if_missing(url_lsoa01_bdy_geojson, lsoa01_geojson)

# Lookups
lsoa01_to11_rds <- file.path(data_dir, "lsoa01_to11_lookup.rds")
lookup_ok <- FALSE
if (file.exists(lsoa01_to11_rds)) {
  existing_lookup <- tryCatch(readRDS(lsoa01_to11_rds), error = function(e) NULL)
  if (!is.null(existing_lookup) && all(c("LSOA01CD", "LSOA11CD", "LAD11CD") %in% toupper(names(existing_lookup)))) {
    lookup_ok <- TRUE
  }
}

if (!lookup_ok) {
  lkp <- fetch_arcgis_attrs(
    base_query_url = url_lsoa01_to_lsoa11_service,
    out_fields = c("LSOA01CD", "LSOA11CD", "LAD11CD", "CHGIND")
  )
  saveRDS(lkp, lsoa01_to11_rds)
}

# Deaths archive
deaths_zip <- file.path(data_dir, "deathsbylsoamidyear01to19.zip")
download_if_missing(ons_deaths_zip_url, deaths_zip)
```

## Read and Harmonize Inputs

```{r assemble-data, include=FALSE}
# Read boundaries
lsoa11_bdy <- st_read(lsoa11_geojson, quiet = TRUE) %>% st_make_valid()
lsoa01_bdy <- st_read(lsoa01_geojson, quiet = TRUE) %>% st_make_valid()

# Read lookups
lsoa01_to11 <- readRDS(lsoa01_to11_rds)

# Standardize lookup names
setnames(lsoa01_to11, old = names(lsoa01_to11), new = toupper(names(lsoa01_to11)))
lsoa01_to11 <- lsoa01_to11 %>%
  mutate(
    LSOA01CD = norm_code(LSOA01CD),
    LSOA11CD = norm_code(LSOA11CD),
    LAD11CD = norm_code(LAD11CD)
  )

# Manchester LSOAs
man_lsoa11 <- lsoa01_to11 %>%
  filter(LAD11CD %in% norm_code(target_lad_codes)) %>%
  distinct(LSOA11CD) %>%
  pull(LSOA11CD)

man_lsoa01 <- lsoa01_to11 %>%
  filter(LAD11CD %in% norm_code(target_lad_codes) | LSOA11CD %in% man_lsoa11) %>%
  distinct(LSOA01CD) %>%
  pull(LSOA01CD)

# Subset boundaries to Manchester footprint
code_col_11 <- pick_first_existing_col(lsoa11_bdy, c("LSOA11CD", "LSOA_CODE", "lsoa11cd", "lsoa11cd"))
code_col_01 <- pick_first_existing_col(lsoa01_bdy, c("LSOA01CD", "LSOA_CODE", "lsoa01cd", "lsoa01cd"))

lsoa11_mcr <- lsoa11_bdy %>%
  mutate(LSOA11CD = norm_code(.data[[code_col_11]])) %>%
  filter(LSOA11CD %in% man_lsoa11) %>%
  st_transform(uk_grid_crs)

lsoa01_mcr <- lsoa01_bdy %>%
  mutate(LSOA01CD = norm_code(.data[[code_col_01]])) %>%
  filter(LSOA01CD %in% man_lsoa01) %>%
  st_transform(uk_grid_crs)

# Read deaths and standardize columns
deaths_long <- read_ons_deaths_table(deaths_zip, years_keep = years)

# Aggregate by year + code (safe if raw table has multiple rows per LSOA-year)
deaths_01 <- deaths_long %>%
  filter(year <= 2009, !is.na(lsoa01cd)) %>%
  group_by(year, lsoa01cd) %>%
  summarise(deaths = sum(deaths), .groups = "drop")

deaths_11 <- deaths_long %>%
  filter(year >= 2010, !is.na(lsoa01cd)) %>%
  left_join(
    lsoa01_to11 %>% distinct(LSOA01CD, LSOA11CD),
    by = c("lsoa01cd" = "LSOA01CD")
  ) %>%
  filter(!is.na(LSOA11CD)) %>%
  group_by(year, lsoa11cd = LSOA11CD) %>%
  summarise(deaths = sum(deaths), .groups = "drop")
```

## WorldPop Under-5 Population Rasters

```{r worldpop-download-and-build, include=FALSE}
# WorldPop 2000-2020 age/sex grids. Under-5 is approximated as age bands 0 and 1 for both sexes.
# Filenames used here follow the Global_2000_2020 convention, e.g. gbr_f_0_2008.tif.

wp_base <- "https://data.worldpop.org/GIS/AgeSex_structures/Global_2000_2020"
ages_u5 <- c(0, 1)
sexes <- c("f", "m")

worldpop_tbl <- expand.grid(
  year = years,
  sex = sexes,
  age = ages_u5,
  stringsAsFactors = FALSE
) %>%
  as_tibble() %>%
  mutate(
    url = sprintf("%s/%s/GBR/gbr_%s_%s_%s.tif", wp_base, year, sex, age, year),
    file = file.path(data_dir, sprintf("worldpop_gbr_%s_%s_%s.tif", sex, age, year))
  )

invisible(mapply(download_if_missing, worldpop_tbl$url, worldpop_tbl$file))

# Build yearly under-5 population rasters by summing age-band rasters, then
# clip/project to Manchester and write file-backed rasters to control memory use.
lsoa11_vect <- terra::vect(lsoa11_mcr)
lsoa11_vect <- terra::project(lsoa11_vect, paste0("EPSG:", uk_grid_crs))
mcr_outline <- terra::aggregate(lsoa11_vect, dissolve = TRUE)
wp_yearly <- lapply(years, function(yr) {
  rr <- terra::rast(worldpop_tbl$file[worldpop_tbl$year == yr])
  r_sum <- terra::app(
    rr,
    sum,
    na.rm = TRUE,
    filename = file.path(data_dir, sprintf("wp_u5_sum_%s.tif", yr)),
    overwrite = TRUE,
    wopt = list(gdal = "COMPRESS=LZW")
  )

  mcr_native <- terra::project(mcr_outline, terra::crs(r_sum))
  buffer_native <- if (terra::is.lonlat(mcr_native)) {
    offset_buffer_m / 111320
  } else {
    offset_buffer_m
  }
  mcr_native_buf <- terra::buffer(mcr_native, width = buffer_native)
  r_crop <- terra::crop(
    r_sum,
    mcr_native_buf,
    filename = file.path(data_dir, sprintf("wp_u5_crop_%s.tif", yr)),
    overwrite = TRUE,
    wopt = list(gdal = "COMPRESS=LZW")
  )

  r_proj <- terra::project(
    r_crop,
    terra::crs(lsoa11_vect),
    method = "bilinear",
    filename = file.path(data_dir, sprintf("wp_u5_proj_%s.tif", yr)),
    overwrite = TRUE,
    wopt = list(gdal = "COMPRESS=LZW")
  )

  r_mask <- terra::mask(
    r_proj,
    mcr_outline,
    filename = file.path(data_dir, sprintf("wp_u5_mask_%s.tif", yr)),
    overwrite = TRUE,
    wopt = list(gdal = "COMPRESS=LZW")
  )

  r_out <- if (!is.null(offset_agg_fact) && offset_agg_fact > 1) {
    terra::aggregate(
      r_mask,
      fact = offset_agg_fact,
      fun = sum,
      na.rm = TRUE,
      filename = file.path(data_dir, sprintf("wp_u5_mask_agg%s_%s.tif", offset_agg_fact, yr)),
      overwrite = TRUE,
      wopt = list(gdal = "COMPRESS=LZW")
    )
  } else {
    r_mask
  }

  # Ensure yearly offsets are explicitly in UK National Grid (same as lsoa11_vect).
  terra::project(
    r_out,
    terra::crs(lsoa11_vect),
    method = "bilinear",
    filename = file.path(data_dir, sprintf("wp_u5_ukgrid_%s.tif", yr)),
    overwrite = TRUE,
    wopt = list(gdal = "COMPRESS=LZW")
  )
})
names(wp_yearly) <- as.character(years)

# Quick cell-count check for memory diagnostics.
sapply(wp_yearly, terra::ncell)
```

```{r plotPop, fig.cap='Population (WorldPop under-5) by year',  fig.subcap = names(wp_yearly), out.width='0.49\\linewidth',  echo=FALSE}

popCol = mapmisc::colourScale(
  unlist(lapply(wp_yearly, terra::minmax)),
  breaks= 9, style='equal', transform='sqrt',
  digits=1, dec=2, col='YlOrRd')
  breaks= 12, style='equal', transform='log',
  dec=1, col='YlOrRd')

for(D in names(wp_yearly)) {
  mapmisc::map.new(wp_yearly[[1]])
  terra::plot(wp_yearly[[D]], breaks=popCol$breaks, col=popCol$col, 
    legend=FALSE, add=TRUE)
}
mapmisc::legendBreaks("bottomleft", popCol, inset=0, bty='n')
```

## Build Time-Varying Polygon + Offset Lists

```{r build-time-varying-objects, include=FALSE}
sanitize_poly <- function(x, id_col = "AREA_ID") {
  x <- sf::st_make_valid(x)
  x <- sf::st_collection_extract(x, "POLYGON", warn = FALSE)
  x <- sf::st_make_valid(x)
  x <- x[!sf::st_is_empty(x), ]
  if (nrow(x) == 0) return(x)
  x %>%
    group_by(.data[[id_col]]) %>%
    summarise(deaths = sum(deaths, na.rm = TRUE), do_union = TRUE, .groups = "drop")
}

poly_list <- list()
offset_list <- list()

for (yr in years) {
  if (yr <= 2009) {
    poly_yr <- lsoa01_mcr %>%
      left_join(deaths_01 %>% filter(year == yr), by = c("LSOA01CD" = "lsoa01cd")) %>%
      mutate(
        AREA_ID = LSOA01CD,
        deaths = replace_na(deaths, 0)
      )
  } else {
    poly_yr <- lsoa11_mcr %>%
      left_join(deaths_11 %>% filter(year == yr), by = c("LSOA11CD" = "lsoa11cd")) %>%
      mutate(
        AREA_ID = LSOA11CD,
        deaths = replace_na(deaths, 0)
      )
  }

  # disaggMultiMap expects sf polygons and one aggregation raster per time slice.
  poly_list[[as.character(yr)]] <- sanitize_poly(poly_yr, id_col = "AREA_ID")
  offset_list[[as.character(yr)]] <- wp_yearly[[as.character(yr)]]
}
```

```{r diagnostics}
# Quick diagnostics
sapply(poly_list, nrow)
sapply(poly_list, function(x) sum(x$deaths, na.rm = TRUE))
sapply(poly_list, function(x) table(x$deaths>0))
```

```{r yearly-input-maps, fig.cap="Yearly observed deaths in Manchester.", fig.subcap=paste("Deaths", years),  out.width='0.49\\linewidth', echo=FALSE}
years_chr <- as.character(years)
```{r yearly-input-maps, fig.cap="Yearly observed deaths in Manchester.", fig.subcap=paste("Deaths", names(poly_list)), fig.show='hold', out.width='0.32\\linewidth', fig.width=4, fig.height=4}
years_chr <- names(poly_list)


bgMap = mapmisc::openmap(poly_list[[1]], buffer=10*1000)

deathsCol = mapmisc::colourScale(
  breaks=seq(0, 
    max(sapply(poly_list, function(x) max(x$deaths, na.rm = TRUE)))
    ),
  col='YlOrRd',style='fixed')
  as.vector(sapply(poly_list, function(x) range(x$deaths, na.rm = TRUE))),
  col='YlOrRd', breaks=9, style='equal',
  transform='sqrt', dec=0, opacity = 0.6)

for (yr in years_chr) {
  p <- poly_list[[yr]]
  pv <- terra::vect(p)
  
  deathsColHere <- mapmisc::colourScale(
    pv$deaths,
    breaks = deathsCol$breaks, col=deathsCol$col, 
    style = "fixed")
    pv$deaths,
    breaks = deathsCol$breaks, col=deathsCol$colOpacity, style = "fixed")

  mapmisc::map.new(pv)
  plot(pv, col = deathsColHere$plot, 
    border = "grey35", lty = 0, add = TRUE)
  plot(bgMap, add=TRUE)
  terra::plot(pv, col = deathsColHere$plot, border = "grey35", lwd = 0.2, add = TRUE)
}

  mapmisc::legendBreaks("bottomright", deathsCol, bty = "n", inset=0)
```

## Prepare `disaggMultiMap` Data Object

```{r prepare-disagg-data, cache=TRUE, include=FALSE}
# Use the first year's bbox to scale mesh defaults.
bb <- st_bbox(poly_list[[1]])
hyp <- sqrt((bb$xmax - bb$xmin)^2 + (bb$ymax - bb$ymin)^2)
maxedge <- hyp / 16

dis_data_mmap <- disaggMultiMap::prepare_data_mmap(
  polygon_shapefile_list = poly_list,
  aggregation_rasters_list = offset_list,
  response_var = "deaths",
  id_var = "AREA_ID",
  na_action = TRUE,
  verbose = TRUE,
  mesh_args = list(
    max.edge = c(maxedge / 5, maxedge),
    cutoff = 0.2 * maxedge,
    offset = c(hyp / 16, hyp / 4),
    resolution = 250
  )
)
```

```{r plotdisdata, eval=FALSE}
plot(dis_data_mmap)
```

## Priors and Model  

```{r priors, include=FALSE}
my_priors <- disaggMultiMap::get_priors(dis_data_mmap)

# Example customizations similar to disagg.Rmd.
my_priors$prior_rho_min <- 2500
my_priors$prior_rho_prob <- 0.5
my_priors$prior_iideffect_sd_max <- 0.1
my_priors$prior_iideffect_sd_prob <- 0.5
```

```{r fit-pooled-nb, cache=TRUE}
fm_nb_pooled <- disaggMultiMap::disag_model_mmap(
  data = dis_data_mmap,
  aghq_k = 2,
  family = "negbinomial",
  link = "log",
  iid = TRUE,
  verbose = TRUE,
  priors = my_priors,
  engine = "AGHQ",
  time_varying_betas = FALSE
)
```

```{r predict-pooled}
pred_pooled <- predict(fm_nb_pooled, verbose = TRUE)
summary(fm_nb_pooled)
```

## Pooled Maps

```{r pooled-map-build, eval=TRUE}
# Build posterior median and exceedance (Pr[log-risk > 0]) rasters
# following the same approach as disagg.Rmd.
samps <- aghq::sample_marginal(fm_nb_pooled$aghq_model, 500)
mats <- disaggMultiMap:::get_predict_matrices(
  data = fm_nb_pooled$data,
  new_data = NULL,
  expected_cov_names = tryCatch(
    fm_nb_pooled$model_setup$coef_meta$cov_names,
    error = function(e) NULL
  ),
  time_varying_betas = isTRUE(fm_nb_pooled$model_setup$time_varying_betas)
)

rfield <- mats$A %*% samps$samps
pred_surface <- terra::rast(
  as.data.frame(cbind(
    mats$coords,
    med = apply(rfield, 1, median),
    exc = apply(rfield, 1, function(xx) mean(xx > 0))
  )),
  crs = terra::crs(terra::vect(lsoa11_mcr))
)

manchester_outline <- terra::aggregate(terra::vect(lsoa11_mcr), dissolve = TRUE)
pred_mask <- terra::mask(pred_surface, manchester_outline)
```

```{r pooled-maps, out.width='0.49\\linewidth', fig.cap='Result maps for Manchester : posterior median (log-risk) and exceedance probability.', fig.subcap = c('est','exc'), echo=FALSE}
med_col <- mapmisc::colourScale(
  pred_mask[["med"]],
  breaks = 8,
  style = "equal",
  col = "Spectral",
  rev = TRUE, digits=1
)
exc_col <- mapmisc::colourScale(
  breaks = c(0, 0.2, 0.5, 0.8, 0.95, 1),
  style = "fixed",
  col = "RdYlGn",
  rev = TRUE
)

mapmisc::map.new(pred_mask)
plot(pred_mask[["med"]], col = med_col$col, breaks = med_col$breaks, legend = FALSE, add = TRUE)
plot(terra::vect(lsoa11_mcr), add = TRUE, border = "grey30", lwd = 0.2)
mapmisc::legendBreaks("bottomleft", med_col, bty = "n")

mapmisc::map.new(pred_mask)
plot(pred_mask[["exc"]], col = exc_col$col, breaks = exc_col$breaks, legend = FALSE, add = TRUE)
plot(terra::vect(lsoa11_mcr), add = TRUE, border = "grey30", lwd = 0.2)
mapmisc::legendBreaks("bottomleft", exc_col, bty = "n")
```
